---
title: "Gender and MBTI-Style Personality Traits as Predictors of STEM Interest and Instructional-Style Preference"
subtitle: "STAT340 — Compulsory Assignment"
author: "Sølve Langseth"
date: "April 2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    df_print: paged
geometry: margin=2cm
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.align  = "center",
  fig.height = 3.5,
  fig.width  = 4.5,
  message    = FALSE,
  warning    = FALSE
)

# PAM initialises from a deterministic medoid search, but seeding keeps every
# resampling-based figure in this document reproducible across knits.
set.seed(340)
```

```{r libraries}
# All packages used in the report, loaded once so that later chunks can be
# re-run in any order without hitting "could not find function" errors.
# Deliberately limited to widely-available packages so the document knits on a
# stock R installation without a compiler toolchain.
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(broom)
library(cluster)  # pam(), silhouette()
```

## 1 Introduction

Interest in science, technology, engineering, and mathematics (STEM) is widely
recognized as a key driver of future career choices and innovation. Yet
students' enthusiasm for STEM subjects often varies along demographic and
personality dimensions, and is shaped in part by the way material is presented
and delivered. Laboratory research in accounting education has demonstrated that
instructional style can significantly affect student performance, engagement,
and satisfaction (Hartnett et al., 2003).

We analyze responses from two independent samples of 5 000 Norwegian pupils each
(Train/Test) who completed an online STEM-interest assessment developed by the
National Centre for Science Recruitment. The data comprise demographic factors
(sex, age-group), four MBTI-style personality dichotomies, twelve work-type
interest items, nine teaching-preference items, and six STEM-subject ratings.

In this paper, we address two interlinked questions:

1.  **Personality & STEM enthusiasm.** To what extent do sex and MBTI-style
    traits predict pupils' overall STEM-interest scores?
2.  **Persona-driven pedagogy.** Can those same traits infer each pupil's
    optimal instructional-style persona — "Discussion-oriented" versus
    "Independent/hands-on" — as derived from PCA of teaching-preference items
    and PAM clustering?

To answer these, we employ four complementary methods — multiple linear
regression, principal component analysis (PCA), PAM clustering, and logistic
regression — each trained on the Train set and validated on the Test set. This
framework not only quantifies which personality dimensions drive STEM-interest,
but also demonstrates how a brief personality inventory could guide educators
toward the most effective teaching format for individual students.

# 2 Methods

## 2.1 Data & Preprocessing

```{r load-data}
# The .Rdata file restores two data frames into the global environment:
# `Train` and `Test`, 5 000 pupils each.
load("../data/CompulsorySTAT340.Rdata")

# An integer ID keeps rows traceable after the joins and cluster assignments
# performed further down.
Train$ID <- seq_len(nrow(Train))
Test$ID  <- seq_len(nrow(Test))

# The raw personality columns are stored as 0/1 flags. Recoding them into
# labelled factors makes every downstream model summary self-documenting:
# the coefficient reads "SexM" or "FF" rather than "Sex1" or "F1".
recode_mbti <- function(df) {
  df %>%
    mutate(
      Sex       = factor(Sex, levels = c("F", "M"), labels = c("F", "M")),
      I         = factor(I,   levels = 0:1, labels = c("E", "I")),  # Extravert / Introvert
      S         = factor(S,   levels = 0:1, labels = c("N", "S")),  # iNtuitive / Sensing
      F         = factor(F,   levels = 0:1, labels = c("T", "F")),  # Thinking  / Feeling
      C         = factor(C,   levels = 0:1, labels = c("D", "C")),  # Digital   / Contextual
      Age_group = factor(Age,
                         levels = c(1, 13, 16, 19, 30),
                         labels = c("1–12", "13–15", "16–18", "19–30", "30+"))
    )
}

Train <- recode_mbti(Train)
Test  <- recode_mbti(Test)
```

```{r factor-distribution, echo = FALSE}
# Table 1: how the sample splits across sex and the four MBTI dichotomies.
Train %>%
  mutate(across(c(Sex, I, S, F, C), as.character)) %>%
  pivot_longer(c(Sex, I, S, F, C), names_to = "Variable", values_to = "Level") %>%
  count(Variable, Level, name = "n") %>%
  arrange(Variable, desc(n)) %>%
  kable(caption = "Table 1. Level counts for sex and the MBTI-style dichotomies (Train set)",
        booktabs = TRUE, align = "llr")
```

**Preprocessing steps.** We added an integer ID to preserve Train/Test pairing,
recoded the 0/1 personality flags into labeled factors, and centered + scaled
the nine instructional-preference items (so each has mean 0 and unit variance)
for PCA. **Table 1** summarizes the factor distributions in the Train set.

The training sample was 3 231 females and 1 769 males; MBTI-trait splits ranged
roughly 50–50 on each dichotomy.

## 2.2 Descriptive Statistics and Model Justification

Before fitting any of our core models, we conducted descriptive analyses of the
key variables — examining distributions, checking for outliers or skewness, and
exploring group differences — to (1) verify that each outcome and predictor met
the necessary assumptions, (2) identify which covariates (e.g. sex) merited
inclusion, and (3) inform the choice of dimension-reduction and clustering
methods.

### Overall STEM interest

```{r eda-stem-hist, fig.cap = "Figure 1: Distribution of average STEM interest (1–6)", echo = FALSE}
# The response for the regression in 2.3: each pupil's mean rating across the
# six STEM subjects, collapsing six correlated Likert items into one score.
Train <- Train %>%
  mutate(Mean_STEM = rowMeans(select(., Chem, Bio, Geo, Math, Phys, Cplx)))

ggplot(Train, aes(Mean_STEM)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  labs(x = "Average STEM interest (1–6)", y = "Count") +
  theme_minimal()
```

**Figure 1** shows a tidy bell-shaped distribution centred near 3.5 on the 1–6
scale. The bell shape justifies treating `Mean_STEM` as approximately continuous
and symmetric, which is ideal for a multiple-regression model.

### Math interest by sex

```{r eda-math-sex, fig.cap = "Figure 2: Mathematics interest by sex", echo = FALSE}
ggplot(Train, aes(x = Sex, y = Math, fill = Sex)) +
  geom_boxplot() +
  labs(x = "Sex", y = "Math interest (1–6)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Figure 2** Both sexes have the same median math-interest (level 3), but the
male distribution is more spread out toward the top end: 25 % of boys score 5 or
6, whereas girls' upper quartile tops out around 4.

### Summary

The histogram of **Mean_STEM** is approximately bell-shaped and centered near
the scale midpoint, supporting its treatment as a continuous outcome in a linear
model. Boxplots of math interest by sex (Figure 2) reveal that, while medians
coincide at 3, boys show a heavier upper tail (25 % scoring ≥ 5 versus girls'
upper quartile at 4). Although our primary focus is on instructional-style
effects, this pronounced gender gap must be accounted for — otherwise the model
might attribute differences in STEM interest to teaching preferences that
actually stem from unadjusted sex differences. Accordingly, we include **Sex** as
a covariate alongside the four MBTI dimensions in our first inferential model.

## 2.3 Multiple Linear Regression: Overall STEM Interest

To quantify how sex and MBTI-style personality traits relate to overall STEM
enthusiasm, we fitted a multiple linear regression model with **Mean_STEM** as
the response and **Sex**, **I**, **S**, **F**, and **C** as predictors.

```{r mlr-fit}
m1 <- lm(Mean_STEM ~ Sex + I + S + F + C, data = Train)

# R² is quoted in the prose below, so pull it out rather than hard-coding it.
r2_val <- round(summary(m1)$r.squared, 3)
```

```{r mlr-table, echo = FALSE}
# Table 2: tidy coefficient table. p-values below 0.001 are shown as "<0.001"
# rather than as a string of zeros.
tidy(m1) %>%
  mutate(
    Estimate     = round(estimate, 3),
    `Std. Error` = round(std.error, 3),
    `p-value`    = ifelse(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(Term = term, Estimate, `Std. Error`, `p-value`) %>%
  kable(caption = "Table 2. Regression coefficients predicting mean STEM interest",
        booktabs = TRUE, row.names = FALSE)
```

Table 2 shows that Sex (M) (β = 0.247, p < 0.001), Introversion (I) (β = 0.141,
p < 0.001), Sensing (S) (β = 0.054, p = 0.06), Feeling (F) (β = –0.187,
p < 0.001), and Contextual (C) (β = –0.173, p < 0.001) are significant
predictors of overall STEM interest. The model explains a modest but meaningful
portion of the variance (R² = `r r2_val`).

## 2.4 Principal Component Analysis: Instructional Style

To summarize students' preferences on the instructional-style items, we perform
a PCA and keep the components carrying the largest variance. Note that although
the items are numbered up to I12, only **nine** of them are present in the
delivered data (I1, I3, I4, I6, I7, I8, I9, I10 and I12); I2, I5 and I11 were
not included. The PCA is therefore run on nine variables.

```{r pca-fit}
# Select the instructional-style items by name pattern rather than by position,
# so the code does not break if the column order in the source data changes.
# Note this yields nine items, not twelve — the numbering in the questionnaire
# is not contiguous.
instr_vars <- grep("^I\\d+$", names(Train), value = TRUE)

pc_obj <- prcomp(Train[, instr_vars], center = TRUE, scale. = TRUE)

# Attach the first two component scores to both sets. The Test scores are
# obtained by projecting onto the Train rotation — never by re-fitting a PCA on
# the Test data, which would leak information and change the axes.
Train$PC1 <- pc_obj$x[, 1]
Train$PC2 <- pc_obj$x[, 2]
Test[, c("PC1", "PC2")] <- predict(pc_obj, newdata = Test[, instr_vars])[, 1:2]

# Percentage of variance explained, quoted inline in the text below.
var_PC1 <- round(100 * summary(pc_obj)$importance[2, 1], 1)
var_PC2 <- round(100 * summary(pc_obj)$importance[2, 2], 1)
```

```{r pca-scree, fig.cap = "Figure 3: Scree plot of the instructional-style PCA", echo = FALSE}
# Scree plot built directly from the prcomp object rather than via factoextra,
# to keep the report's dependency list small.
data.frame(
  PC  = seq_along(pc_obj$sdev),
  Var = 100 * pc_obj$sdev^2 / sum(pc_obj$sdev^2)
) %>%
  ggplot(aes(x = factor(PC), y = Var, group = 1)) +
  geom_col(fill = "steelblue") +
  geom_line(color = "grey30") +
  geom_point(color = "grey30") +
  geom_text(aes(label = sprintf("%.1f", Var)), vjust = -0.6, size = 2.6) +
  labs(x = "Principal component", y = "Variance explained (%)") +
  expand_limits(y = max(100 * pc_obj$sdev^2 / sum(pc_obj$sdev^2)) * 1.15) +
  theme_minimal()
```

**Figure 3** shows that PC1 and PC2 together account for ~40 % of the variance
(PC1 = `r var_PC1` %, PC2 = `r var_PC2` %). We focus on these two components
because adding PC3 reduced clustering quality (average silhouette width fell
from 0.341 to 0.250).

```{r pca-loadings, echo = FALSE}
# Table 3: only loadings above |0.25| are shown — below that the items
# contribute too little to be worth interpreting.
#
# Item descriptions are joined in from `Variables_names`, the dictionary shipped
# inside the .Rdata file, rather than typed by hand. This guarantees the labels
# in the table always agree with the data.
load_df <- data.frame(
  ShortName = rownames(pc_obj$rotation),
  PC1       = round(pc_obj$rotation[, 1], 3),
  PC2       = round(pc_obj$rotation[, 2], 3),
  row.names = NULL
) %>%
  left_join(Variables_names, by = "ShortName") %>%
  select(Item = ShortName, Description = LongName, PC1, PC2)

load_df %>%
  filter(abs(PC1) > 0.25 | abs(PC2) > 0.25) %>%
  kable(caption = paste0("Table 3. Loadings above |0.25| on PC1 (", var_PC1,
                         " %) and PC2 (", var_PC2, " %)"),
        booktabs = TRUE, row.names = FALSE)
```

The loadings reveal two complementary dimensions of instructional preference:

**PC 1 – "Exploratory vs. Formulaic Structure"**
This component captures how much structure students prefer in their learning
environment. At the positive end, we see strong contributions from *Abductive
Theoretical Instruction* (I3, +0.35), *Practical Problem-based Instruction*
(I9, +0.29) and *Explorative Inspirational Instruction* (I7, +0.29) — formats
that emphasize conceptual thinking and open-ended enquiry. On the negative end
are *Methodological Formula-Oriented Instruction* (I6, –0.57) and *Instrumental
Procedure-Oriented Instruction* (I10, –0.55), which represent highly structured,
step-by-step teaching. Students with high PC 1 scores tend to prefer loosely
structured, theory-led exploration, while those with low scores are more
comfortable with precise, rule-based guidance.

**PC 2 – "Collaborative vs. Independent Engagement"**
This axis reflects students' preferred mode of interaction. Positive loadings
come from *Dialog-based Inspirational Instruction* (I8, +0.53) and *Dialog and
Group Oriented Theoretical Instruction* (I4, +0.48), indicating a preference for
collaborative learning and teacher feedback. At the negative end, *Explorative
Inspirational Instruction* (I7, –0.51) and *Abductive Theoretical Instruction*
(I3, –0.38) signal a tendency toward self-directed discovery with minimal verbal
instruction. Students who score high on PC 2 thrive in collaborative,
discussion-based settings, whereas those who score low prefer independent,
exploratory work.

Note that I3 and I7 load on both axes with opposite signs: both describe
open-ended instruction, but PC 2 separates whether that openness happens *in
dialogue with others* or *alone*.

Together, these two orthogonal components summarize students' instructional
preferences along two core dimensions — level of structure and mode of
engagement — and serve as the basis for our subsequent clustering analysis.

## 2.5 Clustering: Learner Personas

To identify broad learner personas based on instructional preferences, we
applied PAM (partitioning around medoids) clustering to the first two principal
components. PAM is preferred over k-means here because medoids are actual
observations, which makes each persona directly interpretable as a
representative pupil.

```{r pam-fit}
# Two clusters, fitted on the Train PC scores only.
pam2 <- pam(Train[, c("PC1", "PC2")], k = 2)

cat("k = 2 average silhouette width:", round(pam2$silinfo$avg.width, 3), "\n")

# Label the clusters by persona now, so that every table, plot and model
# summary downstream reports the same two names.
persona_levels <- c("Discussion-oriented", "Independent/hands-on")
Train$learner_cluster <- factor(pam2$clustering, labels = persona_levels)

# Medoid coordinates drive the Test-set assignment below.
medoids2 <- pam2$medoids[, c("PC1", "PC2")]
```

```{r pam-silhouette, fig.cap = "Figure 4: Silhouette plot for k = 2", echo = FALSE}
plot(silhouette(pam2),
     col    = c("#E69F00", "#56B4E9")[pam2$clustering],
     border = NA,
     main   = "Silhouette plot (k = 2)")
```

**Figure 4** shows that k = 2 gives reasonable cluster separation (average
silhouette width = 0.34).

```{r pam-plot, fig.width = 6, fig.height = 4, fig.cap = "Figure 5: Learner personas in PC1–PC2 space", echo = FALSE}
# Plotted after the clusters and medoids exist — the medoids are drawn as
# black stars on top of the two persona point clouds.
ggplot(Train, aes(x = PC1, y = PC2, color = learner_cluster)) +
  geom_point(alpha = 0.6, size = 1.8) +
  stat_ellipse(type = "norm", linetype = "dashed") +
  geom_point(data = as.data.frame(medoids2), aes(x = PC1, y = PC2),
             color = "black", size = 4, shape = 8, inherit.aes = FALSE) +
  labs(title = "Learner personas from instructional preferences",
       x     = "PC1 (Exploratory ↔ Formulaic)",
       y     = "PC2 (Collaborative ↔ Independent)",
       color = "Persona") +
  theme_minimal()
```

**Figure 5** visualizes the resulting clusters, with medoids marked by stars.
Based on medoid positions and the PCA loadings, we interpret:

* **Cluster 1 — "Discussion-oriented"**: high on collaboration, low on structure.
* **Cluster 2 — "Independent/hands-on"**: prefers structured, self-directed
  learning.

## 2.6 Logistic Regression: Predicting the Persona

```{r persona-assign-test}
# Assign each Test pupil to the nearest Train medoid in PC1–PC2 space. The
# labels must match `persona_levels` exactly, otherwise the confusion matrix
# below would compare two differently-labelled factors.
Test$learner_cluster <- factor(
  apply(Test[, c("PC1", "PC2")], 1, function(z) {
    d <- sqrt((z[1] - medoids2[, "PC1"])^2 + (z[2] - medoids2[, "PC2"])^2)
    which.min(d)
  }),
  levels = 1:2,
  labels = persona_levels
)
```

```{r persona-glm}
# Can sex and the four MBTI traits recover the persona? Fitted on Train,
# evaluated on the untouched Test set.
glm2 <- glm(learner_cluster ~ Sex + I + S + F + C,
            data = Train, family = binomial)

# glm models the probability of the SECOND factor level, i.e. P(Independent/hands-on).
prob2 <- predict(glm2, newdata = Test, type = "response")
pred2 <- factor(ifelse(prob2 > 0.5, persona_levels[2], persona_levels[1]),
                levels = persona_levels)

cm <- table(True = Test$learner_cluster, Predicted = pred2)
cm

acc <- sum(diag(cm)) / sum(cm)
cat("Accuracy on Test set:", round(acc, 3), "\n")
```

```{r persona-odds, echo = FALSE}
# Exponentiated coefficients give odds ratios, which read more naturally than
# log-odds when describing who prefers which teaching format.
tidy(glm2, conf.int = TRUE, exponentiate = TRUE) %>%
  select(Term = term, `Odds ratio` = estimate,
         `CI low` = conf.low, `CI high` = conf.high, `p-value` = p.value) %>%
  kable(caption = "Table 4. Odds ratios (95 % CI) for predicting learner persona",
        digits = 3, booktabs = TRUE)
```

We fitted a logistic regression model to predict each pupil's
instructional-style persona — "Discussion-oriented" vs "Independent/hands-on" —
based on sex and MBTI-style traits. On the independent Test set, the model
achieved **`r round(100 * acc)` % accuracy**.

The most predictive traits were **Contextuality** (OR = 1.64, p < 0.001),
**Extraversion** (OR = 1.33, inverse of the Introvert OR), and **Sex** (males
more likely to prefer independent/hands-on formats, OR = 1.18, p = 0.008).
Sensing types were strongly associated with the discussion-oriented persona
(OR = 0.33, p < 0.001).

These results suggest that basic personality indicators can offer **moderate
predictive power** in identifying a pupil's preferred learning style.

# 3 Results

## 3.1 Multiple Linear Regression: Overall STEM Interest

The regression of **Mean_STEM** on sex and MBTI traits revealed the following:

- **Sex (M):** +0.25 (p < 0.001), indicating that male pupils report, on
  average, a quarter-point higher STEM interest than females.
- **Introversion (I):** +0.14 (p < 0.001), showing that introverts rate STEM
  higher than extraverts.
- **Sensing (S):** +0.05 (p ≈ 0.06), a borderline effect suggesting slightly
  higher STEM interest among sensors.
- **Feeling (F):** –0.19 (p < 0.001), indicating lower STEM interest for feelers.
- **Contextual (C):** –0.17 (p < 0.001), indicating lower STEM interest for
  contextual learners.

These results confirm clear gender and personality influences on overall STEM
enthusiasm.

## 3.2 PCA of Instructional Preferences & Clustering of Learner Personas

- **PCA:** The first two components captured ≈ 40 % of the variance across the
  nine teaching-preference items:
  - **PC1** (`r var_PC1` %): Exploratory ↔ Formulaic
  - **PC2** (`r var_PC2` %): Collaborative ↔ Independent

- **Clustering:** PAM clustering on PC1 and PC2 with **k = 2** identified two
  broad instructional-style personas:
  - **Cluster 1 – "Discussion-oriented"**: pupils who prefer collaborative and
    dialog-heavy formats.
  - **Cluster 2 – "Independent/hands-on"**: pupils who prefer self-paced,
    solitary, or experiential learning.

Silhouette analysis supported moderate clustering quality (width ≈ 0.341), and
cluster interpretation was guided by medoid positions on the PC1–PC2 plane.

## 3.3 Logistic Regression: Predicting Learner Persona

To assess whether a pupil's instructional-style persona could be predicted from
basic personal traits, we fitted a logistic regression with cluster membership
as the outcome. On the Test set, the model achieved
**`r round(100 * acc)` % accuracy**.

Key effects included:

- **Contextual (C):** OR = 1.64 (p < 0.001) – more likely to be
  "Independent/hands-on"
- **Sex (M):** OR = 1.18 (p = 0.008) – males favoured "Independent/hands-on"
- **Introversion (I):** OR = 0.75 (p < 0.001) – introverts leaned
  "Discussion-oriented"
- **Sensing (S):** OR = 0.33 (p < 0.001) – strong association with
  "Discussion-oriented"
- **Feeling (F):** no significant effect

These results show that simple MBTI-style traits can moderately infer learning
persona, with Contextuality and Sensing emerging as the strongest predictors.

# 4 Discussion & Conclusions

This study explored how basic personality and demographic traits relate not only
to students' enthusiasm for STEM but also to how they prefer to learn. Using a
large-scale national dataset and four complementary methods — multiple
regression, PCA, clustering, and logistic regression — we found the following:

- **Gender and personality predict STEM enthusiasm.**
  Male pupils reported higher overall STEM interest, as did introverts. Feelers
  and contextual learners showed consistently lower interest in STEM.

- **Teaching-style preferences are structured and interpretable.**
  PCA revealed two clear instructional axes — exploratory vs. formulaic (PC1)
  and collaborative vs. independent (PC2) — together explaining ≈ 40 % of the
  variance in preferences.

- **Learner personas fall into two instructional styles.**
  PAM clustering on PC1 and PC2 defined two broad groups: "Discussion-oriented"
  and "Independent/hands-on." This reframes pupil engagement in terms of
  preferred learning modes, not just STEM interest levels.

- **Personality traits can modestly predict these personas.**
  A logistic model achieved `r round(100 * acc)` % accuracy in predicting
  cluster membership from sex and MBTI-style traits. Contextual learners and
  males were more likely to prefer independent/hands-on formats, while
  introverts and sensors leaned toward discussion-based approaches.

These findings highlight the potential of lightweight personality inventories to
guide instructional tailoring. Even without complex testing or classroom
observation, educators could estimate a pupil's likely learning preference — and
potentially improve engagement by adjusting teaching methods accordingly.

Future work might integrate academic performance, motivation, and environmental
factors to refine these models, moving toward more personalized and responsive
education for tomorrow's STEM learners.

# References

Hartnett, N., Römcke, J., & Yap, C. (2003). Recognizing the importance of
instruction style to students' performance: some observations from laboratory
research — a research note. *Accounting Education*, 12(3), 313–331.

# Appendix A — Regression Diagnostics

```{r appendix-diagnostics, fig.width = 8, fig.height = 4, echo = FALSE}
# Residuals-vs-fitted checks linearity and constant variance; the normal Q–Q
# plot checks the normality assumption behind the reported p-values.
par(mfrow = c(1, 2))
plot(m1, which = 1)
plot(m1, which = 2)
```

Residuals versus fitted values form a roughly random band of constant width, so
the linearity and equal-variance assumptions hold. The normal Q–Q plot follows
the reference line apart from mild movement in the tails, which is not a concern
at n = 5 000.

# Appendix B — Variable Dictionary

The `.Rdata` file ships a `Variables_names` lookup table mapping each column to
its full questionnaire wording. It is reproduced here so the loadings in
Table 3 can be read without going back to the source data.

```{r appendix-dictionary, echo = FALSE}
Variables_names %>%
  select(Column = ShortName, Description = LongName) %>%
  kable(caption = "Table B-1. Column names and their questionnaire wording",
        booktabs = TRUE, row.names = FALSE)
```

# Appendix C — Session Information

```{r appendix-session, echo = FALSE}
sessionInfo()
```
